---
output:
  workflowr::wflow_html:
    toc: no
    
---

<div>
# Xenium Data Analysis (Without Xenium Ranger){.tabset .tabset-fade}

## Introduction

## Methods

## Example data


</div>

<hr>

# Overview

This report documents the standard single-cell data analysis of a 10x Genomics
Xenium 5K dataset. Sections cover data loading, quality control, normalisation,
dimensionality reduction, clustering, differential expression, and pathway
analysis. All outputs are written to `r params$out_dir`.

!Summary table! 

```{r}
suppressPackageStartupMessages({
  library(Seurat)
  library(SeuratObject)
  library(tidyverse)
  library(RColorBrewer)
  library(viridis)
  library(knitr)
  library(kableExtra)

})

# if (params$run_pathway) {
#   suppressPackageStartupMessages({
#     library(clusterProfiler)
#     library(enrichplot)
#     library(org.Hs.eg.db)
#   })
# }
theme_set(theme_classic(base_size = 12))

```


## Data Loading

### Xenium Data Import

```{r load-data}
# load xenium data using custom script and save processed files. 
out_dir = "../data/GSE307895_processed/"
#meta <- GEOquery::getGEO("GSE307895")[[1]]@phenoData@data
geo = "GSM9233671"
dat <- load_xenium(
  sample_id = "output-220149",
  data.folder = "../data/Raw_data_files",
  mtx.dir = "cell_feature_matrix/matrix.mtx.gz",
  barcode.dir = "cell_feature_matrix/barcodes.tsv.gz",
  features.dir = "cell_feature_matrix/features.tsv.gz",
  cell.data = "cells.csv.gz",
  cell.data.barcodes = "cell_id",
  additional_metadata = c("geo_accession" = geo,
                          "shedding" = meta$`active hsv-2 shedding:ch1`[meta$geo_accession==geo],
                          "follicular" = meta$`folicular phase:ch1`[meta$geo_accession==geo],
                          "status" = meta$`hsv-2 status:ch1`[meta$geo_accession==geo],
                          "sample_id" = meta$`patient id:ch1`[meta$geo_accession==geo], 
                          "nugent" = meta$`nugent result:ch1`[meta$geo_accession==geo], 
                          "nugentscore" = meta$`nugent score:ch1`[meta$geo_accession==geo], 
                          "leuteul" = meta$`leuteul phase:ch1`[meta$geo_accession==geo]),
  coordinates = c("x_centroid", "y_centroid"))
saveRDS(dat, paste0(out_dir,geo,".rds"))

all <- dat 

geo = "GSM9233670"
dat <- load_xenium(
  sample_id = "output-223439",
  data.folder = "../data/Raw_data_files",
  mtx.dir = "cell_feature_matrix/matrix.mtx.gz",
  barcode.dir = "cell_feature_matrix/barcodes.tsv.gz",
  features.dir = "cell_feature_matrix/features.tsv.gz",
  cell.data = "cells.csv.gz",
  cell.data.barcodes = "cell_id",
  additional_metadata = c("geo_accession" = geo,
                          "shedding" = meta$`active hsv-2 shedding:ch1`[meta$geo_accession==geo],
                          "follicular" = meta$`folicular phase:ch1`[meta$geo_accession==geo],
                          "status" = meta$`hsv-2 status:ch1`[meta$geo_accession==geo],
                          "sample_id" = meta$`patient id:ch1`[meta$geo_accession==geo], 
                          "nugent" = meta$`nugent result:ch1`[meta$geo_accession==geo], 
                          "nugentscore" = meta$`nugent score:ch1`[meta$geo_accession==geo], 
                          "leuteul" = meta$`leuteul phase:ch1`[meta$geo_accession==geo]),  coordinates = c("x_centroid", "y_centroid"))
saveRDS(dat, paste0(out_dir,geo,".rds"))
all <- merge(all, dat, merge.dr = T)

geo = "GSM9233669"
dat <- load_xenium(
  sample_id = "output-240679-2",
  data.folder = "../data/Raw_data_files",
  mtx.dir = "cell_feature_matrix/matrix.mtx.gz",
  barcode.dir = "cell_feature_matrix/barcodes.tsv.gz",
  features.dir = "cell_feature_matrix/features.tsv.gz",
  cell.data = "cells.csv.gz",
  cell.data.barcodes = "cell_id",
  additional_metadata = c("geo_accession" = geo,
                          "shedding" = meta$`active hsv-2 shedding:ch1`[meta$geo_accession==geo],
                          "follicular" = meta$`folicular phase:ch1`[meta$geo_accession==geo],
                          "status" = meta$`hsv-2 status:ch1`[meta$geo_accession==geo],
                          "sample_id" = meta$`patient id:ch1`[meta$geo_accession==geo], 
                          "nugent" = meta$`nugent result:ch1`[meta$geo_accession==geo], 
                          "nugentscore" = meta$`nugent score:ch1`[meta$geo_accession==geo], 
                          "leuteul" = meta$`leuteul phase:ch1`[meta$geo_accession==geo]),  coordinates = c("x_centroid", "y_centroid"))
saveRDS(dat, paste0(out_dir,geo,".rds"))
all <- merge(all, dat, merge.dr = T)

geo = "GSM9233668"
dat <- load_xenium(
  sample_id = "output-240679-1",
  data.folder = "../data/Raw_data_files",
  mtx.dir = "cell_feature_matrix/matrix.mtx.gz",
  barcode.dir = "cell_feature_matrix/barcodes.tsv.gz",
  features.dir = "cell_feature_matrix/features.tsv.gz",
  cell.data = "cells.csv.gz",
  cell.data.barcodes = "cell_id",
  additional_metadata = c("geo_accession" = geo,
                          "shedding" = meta$`active hsv-2 shedding:ch1`[meta$geo_accession==geo],
                          "follicular" = meta$`folicular phase:ch1`[meta$geo_accession==geo],
                          "status" = meta$`hsv-2 status:ch1`[meta$geo_accession==geo],
                          "sample_id" = meta$`patient id:ch1`[meta$geo_accession==geo], 
                          "nugent" = meta$`nugent result:ch1`[meta$geo_accession==geo], 
                          "nugentscore" = meta$`nugent score:ch1`[meta$geo_accession==geo], 
                          "leuteul" = meta$`leuteul phase:ch1`[meta$geo_accession==geo]),  coordinates = c("x_centroid", "y_centroid"))
saveRDS(dat, paste0(out_dir,geo,".rds"))
all <- merge(all, dat, merge.dr = T)

geo = "GSM9233667"
dat <- load_xenium(
  sample_id = "output-252499",
  data.folder = "../data/Raw_data_files",
  mtx.dir = "cell_feature_matrix/matrix.mtx.gz",
  barcode.dir = "cell_feature_matrix/barcodes.tsv.gz",
  features.dir = "cell_feature_matrix/features.tsv.gz",
  cell.data = "cells.csv.gz",
  cell.data.barcodes = "cell_id",
  additional_metadata = c("geo_accession" = geo,
                          "shedding" = meta$`active hsv-2 shedding:ch1`[meta$geo_accession==geo],
                          "follicular" = meta$`folicular phase:ch1`[meta$geo_accession==geo],
                          "status" = meta$`hsv-2 status:ch1`[meta$geo_accession==geo],
                          "sample_id" = meta$`patient id:ch1`[meta$geo_accession==geo], 
                          "nugent" = meta$`nugent result:ch1`[meta$geo_accession==geo], 
                          "nugentscore" = meta$`nugent score:ch1`[meta$geo_accession==geo], 
                          "leuteul" = meta$`leuteul phase:ch1`[meta$geo_accession==geo]),  coordinates = c("x_centroid", "y_centroid"))
saveRDS(dat, paste0(out_dir,geo,".rds"))
all <- merge(all, dat, merge.dr = T)

geo = "GSM9233666"
dat <- load_xenium(
  sample_id = "output-233869",
  data.folder = "../data/Raw_data_files",
  mtx.dir = "cell_feature_matrix/matrix.mtx.gz",
  barcode.dir = "cell_feature_matrix/barcodes.tsv.gz",
  features.dir = "cell_feature_matrix/features.tsv.gz",
  cell.data = "cells.csv.gz",
  cell.data.barcodes = "cell_id",
  additional_metadata = c("geo_accession" = geo,
                          "shedding" = meta$`active hsv-2 shedding:ch1`[meta$geo_accession==geo],
                          "follicular" = meta$`folicular phase:ch1`[meta$geo_accession==geo],
                          "status" = meta$`hsv-2 status:ch1`[meta$geo_accession==geo],
                          "sample_id" = meta$`patient id:ch1`[meta$geo_accession==geo], 
                          "nugent" = meta$`nugent result:ch1`[meta$geo_accession==geo], 
                          "nugentscore" = meta$`nugent score:ch1`[meta$geo_accession==geo], 
                          "leuteul" = meta$`leuteul phase:ch1`[meta$geo_accession==geo]),  coordinates = c("x_centroid", "y_centroid"))
saveRDS(dat, paste0(out_dir,geo,".rds"))
all <- merge(all, dat, merge.dr = T)

geo = "GSM9233665"
dat <- load_xenium(
  sample_id = "output-253109",
  data.folder = "../data/Raw_data_files",
  mtx.dir = "cell_feature_matrix/matrix.mtx.gz",
  barcode.dir = "cell_feature_matrix/barcodes.tsv.gz",
  features.dir = "cell_feature_matrix/features.tsv.gz",
  cell.data = "cells.csv.gz",
  cell.data.barcodes = "cell_id",
  additional_metadata = c("geo_accession" = geo,
                          "shedding" = meta$`active hsv-2 shedding:ch1`[meta$geo_accession==geo],
                          "follicular" = meta$`folicular phase:ch1`[meta$geo_accession==geo],
                          "status" = meta$`hsv-2 status:ch1`[meta$geo_accession==geo],
                          "sample_id" = meta$`patient id:ch1`[meta$geo_accession==geo], 
                          "nugent" = meta$`nugent result:ch1`[meta$geo_accession==geo], 
                          "nugentscore" = meta$`nugent score:ch1`[meta$geo_accession==geo], 
                          "leuteul" = meta$`leuteul phase:ch1`[meta$geo_accession==geo]),  coordinates = c("x_centroid", "y_centroid"))
saveRDS(dat, paste0(out_dir,geo,".rds")) # ISSUE?
all <- merge(all, dat, merge.dr = T)

geo = "GSM9233664"
dat <- load_xenium(
  sample_id = "output-220409-2",
  data.folder = "../data/Raw_data_files",
  mtx.dir = "cell_feature_matrix/matrix.mtx.gz",
  barcode.dir = "cell_feature_matrix/barcodes.tsv.gz",
  features.dir = "cell_feature_matrix/features.tsv.gz",
  cell.data = "cells.csv.gz",
  cell.data.barcodes = "cell_id",
  additional_metadata = c("geo_accession" = geo,
                          "shedding" = meta$`active hsv-2 shedding:ch1`[meta$geo_accession==geo],
                          "follicular" = meta$`folicular phase:ch1`[meta$geo_accession==geo],
                          "status" = meta$`hsv-2 status:ch1`[meta$geo_accession==geo],
                          "sample_id" = meta$`patient id:ch1`[meta$geo_accession==geo], 
                          "nugent" = meta$`nugent result:ch1`[meta$geo_accession==geo], 
                          "nugentscore" = meta$`nugent score:ch1`[meta$geo_accession==geo], 
                          "leuteul" = meta$`leuteul phase:ch1`[meta$geo_accession==geo]),  coordinates = c("x_centroid", "y_centroid"))
saveRDS(dat, paste0(out_dir,geo,".rds"))
all <- merge(all, dat, merge.dr = T)

geo = "GSM9233663"
dat <- load_xenium(
  sample_id = "output-220409-1",
  data.folder = "../data/Raw_data_files",
  mtx.dir = "cell_feature_matrix/matrix.mtx.gz",
  barcode.dir = "cell_feature_matrix/barcodes.tsv.gz",
  features.dir = "cell_feature_matrix/features.tsv.gz",
  cell.data = "cells.csv.gz",
  cell.data.barcodes = "cell_id",
  additional_metadata = c("geo_accession" = geo,
                          "shedding" = meta$`active hsv-2 shedding:ch1`[meta$geo_accession==geo],
                          "follicular" = meta$`folicular phase:ch1`[meta$geo_accession==geo],
                          "status" = meta$`hsv-2 status:ch1`[meta$geo_accession==geo],
                          "sample_id" = meta$`patient id:ch1`[meta$geo_accession==geo], 
                          "nugent" = meta$`nugent result:ch1`[meta$geo_accession==geo], 
                          "nugentscore" = meta$`nugent score:ch1`[meta$geo_accession==geo], 
                          "leuteul" = meta$`leuteul phase:ch1`[meta$geo_accession==geo]),  coordinates = c("x_centroid", "y_centroid"))
saveRDS(dat, paste0(out_dir,geo,".rds"))
all <- merge(all, dat, merge.dr = T)

geo = "GSM9233662"
dat <- load_xenium(
  sample_id = "output-222439",
  data.folder = "../data/Raw_data_files",
  mtx.dir = "cell_feature_matrix/matrix.mtx.gz",
  barcode.dir = "cell_feature_matrix/barcodes.tsv.gz",
  features.dir = "cell_feature_matrix/features.tsv.gz",
  cell.data = "cells.csv.gz",
  cell.data.barcodes = "cell_id",
  additional_metadata = c("geo_accession" = geo,
                          "shedding" = meta$`active hsv-2 shedding:ch1`[meta$geo_accession==geo],
                          "follicular" = meta$`folicular phase:ch1`[meta$geo_accession==geo],
                          "status" = meta$`hsv-2 status:ch1`[meta$geo_accession==geo],
                          "sample_id" = meta$`patient id:ch1`[meta$geo_accession==geo], 
                          "nugent" = meta$`nugent result:ch1`[meta$geo_accession==geo], 
                          "nugentscore" = meta$`nugent score:ch1`[meta$geo_accession==geo], 
                          "leuteul" = meta$`leuteul phase:ch1`[meta$geo_accession==geo]),  coordinates = c("x_centroid", "y_centroid"))
saveRDS(dat, paste0(out_dir,geo,".rds"))
all <- merge(all, dat, merge.dr = T)

geo = "GSM9233661"
dat <- load_xenium(
  sample_id = "output-227129",
  data.folder = "../data/Raw_data_files",
  mtx.dir = "cell_feature_matrix/matrix.mtx.gz",
  barcode.dir = "cell_feature_matrix/barcodes.tsv.gz",
  features.dir = "cell_feature_matrix/features.tsv.gz",
  cell.data = "cells.csv.gz",
  cell.data.barcodes = "cell_id",
  additional_metadata = c("geo_accession" = geo,
                          "shedding" = meta$`active hsv-2 shedding:ch1`[meta$geo_accession==geo],
                          "follicular" = meta$`folicular phase:ch1`[meta$geo_accession==geo],
                          "status" = meta$`hsv-2 status:ch1`[meta$geo_accession==geo],
                          "sample_id" = meta$`patient id:ch1`[meta$geo_accession==geo], 
                          "nugent" = meta$`nugent result:ch1`[meta$geo_accession==geo], 
                          "nugentscore" = meta$`nugent score:ch1`[meta$geo_accession==geo], 
                          "leuteul" = meta$`leuteul phase:ch1`[meta$geo_accession==geo]),  coordinates = c("x_centroid", "y_centroid"))
saveRDS(dat, paste0(out_dir,geo,".rds"))
all <- merge(all, dat, merge.dr = T)

geo = "GSM9233660"
dat <- load_xenium(
  sample_id = "output-234919",
  data.folder = "../data/Raw_data_files",
  mtx.dir = "cell_feature_matrix/matrix.mtx.gz",
  barcode.dir = "cell_feature_matrix/barcodes.tsv.gz",
  features.dir = "cell_feature_matrix/features.tsv.gz",
  cell.data = "cells.csv.gz",
  cell.data.barcodes = "cell_id",
  additional_metadata = c("geo_accession" = geo,
                          "shedding" = meta$`active hsv-2 shedding:ch1`[meta$geo_accession==geo],
                          "follicular" = meta$`folicular phase:ch1`[meta$geo_accession==geo],
                          "status" = meta$`hsv-2 status:ch1`[meta$geo_accession==geo],
                          "sample_id" = meta$`patient id:ch1`[meta$geo_accession==geo], 
                          "nugent" = meta$`nugent result:ch1`[meta$geo_accession==geo], 
                          "nugentscore" = meta$`nugent score:ch1`[meta$geo_accession==geo], 
                          "leuteul" = meta$`leuteul phase:ch1`[meta$geo_accession==geo]),  coordinates = c("x_centroid", "y_centroid"))
saveRDS(dat, paste0(out_dir,geo,".rds"))
all <- merge(all, dat, merge.dr = T)

geo = "GSM9233659"
dat <- load_xenium(
  sample_id = "output-237329",
  data.folder = "../data/Raw_data_files",
  mtx.dir = "cell_feature_matrix/matrix.mtx.gz",
  barcode.dir = "cell_feature_matrix/barcodes.tsv.gz",
  features.dir = "cell_feature_matrix/features.tsv.gz",
  cell.data = "cells.csv.gz",
  cell.data.barcodes = "cell_id",
  additional_metadata = c("geo_accession" = geo,
                          "shedding" = meta$`active hsv-2 shedding:ch1`[meta$geo_accession==geo],
                          "follicular" = meta$`folicular phase:ch1`[meta$geo_accession==geo],
                          "status" = meta$`hsv-2 status:ch1`[meta$geo_accession==geo],
                          "sample_id" = meta$`patient id:ch1`[meta$geo_accession==geo], 
                          "nugent" = meta$`nugent result:ch1`[meta$geo_accession==geo], 
                          "nugentscore" = meta$`nugent score:ch1`[meta$geo_accession==geo], 
                          "leuteul" = meta$`leuteul phase:ch1`[meta$geo_accession==geo]),  coordinates = c("x_centroid", "y_centroid"))
saveRDS(dat, paste0(out_dir,geo,".rds"))
all <- merge(all, dat, merge.dr = T)

geo = "GSM9233658"
dat <- load_xenium(
  sample_id = "output-246849",
  data.folder = "../data/Raw_data_files",
  mtx.dir = "cell_feature_matrix/matrix.mtx.gz",
  barcode.dir = "cell_feature_matrix/barcodes.tsv.gz",
  features.dir = "cell_feature_matrix/features.tsv.gz",
  cell.data = "cells.csv.gz",
  cell.data.barcodes = "cell_id",
  additional_metadata = c("geo_accession" = geo,
                          "shedding" = meta$`active hsv-2 shedding:ch1`[meta$geo_accession==geo],
                          "follicular" = meta$`folicular phase:ch1`[meta$geo_accession==geo],
                          "status" = meta$`hsv-2 status:ch1`[meta$geo_accession==geo],
                          "sample_id" = meta$`patient id:ch1`[meta$geo_accession==geo], 
                          "nugent" = meta$`nugent result:ch1`[meta$geo_accession==geo], 
                          "nugentscore" = meta$`nugent score:ch1`[meta$geo_accession==geo], 
                          "leuteul" = meta$`leuteul phase:ch1`[meta$geo_accession==geo]),  coordinates = c("x_centroid", "y_centroid"))
saveRDS(dat, paste0(out_dir,geo,".rds"))
all <- merge(all, dat, merge.dr = T)

saveRDS(all, paste0(out_dir,"all.rds"))

```


```{r}
all <- readRDS(paste0(out_dir,"all.rds"))
FeatureScatter(all, "nCount_Xenium", "nFeature_Xenium", pt.size=2)
table(all$geo_accession, all$nCount_Xenium<10)
all <- subset(all, subset = nCount_Xenium > 10 & cell_area >30)

d1 <- subset(all, subset = geo_accession == "GSM9233671")
d2 <- subset(all, subset = geo_accession == "GSM9233670")

process_data <- function(dat, nfeatures=300, dims=1:20, resolution= 0.3, loudly=F){
  cat("\nNormalizing Data...\n")
  dat <- NormalizeData(dat, verbose=loudly)
    cat("\nFinding Variable Features...\n")
  dat <- FindVariableFeatures(dat, nfeatures = nfeatures, verbose=loudly)
    cat("\nScaling Data...\n")

  dat <- ScaleData(dat, verbose=loudly)
  cat("\nRunning PCA...\n")
  dat <- RunPCA(dat, verbose=loudly)
  cat("\nRunning UMAP...\n")
  dat <- RunUMAP(dat, verbose=loudly, dims = dims)
  cat("\nFinding Neighbours...\n")
  dat <- FindNeighbors(dat, dims = dims, verbose=loudly)
  cat("\nFinding Clusters...\n")
  dat <- FindClusters(dat, resolution = resolution, verbose=loudly)
  return(dat)
}

d1 <- process_data(dat = d1)
d2 <- process_data(dat = d2)  

UMAPPlot(d1, label=T)
UMAPPlot(d2, label=T)

VlnPlot(d1, "PECAM1", pt.size=0)

m <- FindMarkers(d2, ident.1 = 1, only.pos=T)
d1$ref = Idents(d1) #1 and 6 are CD3
d2$ref = Idents(d2) #5 = CD8, 9 = CD8-T cell
```

```{r}
d <- merge(d1,d2, merge.dr = T)
d <- JoinLayers(d)
d[["Xenium"]] <- split(d[["Xenium"]], f = d$geo_accession)

d<-process_data(d, loudly=T)

UMAPPlot(d, label=T, split.by = 'geo_accession', group.by='ref')

d <- IntegrateLayers(object = d, method = RPCAIntegration, orig.reduction = "pca", new.reduction = "integrated.rpca",verbose = T)

d <- JoinLayers(d)
d<- RunUMAP(d, reduction = "integrated.rpca", dims=1:30)

UMAPPlot(d, label=T, split.by = 'geo_accession', group.by='ref')

```

# integrate all data

```{r}
all <- readRDS("../data/GSE307895_processed/all.rds")
all <- subset(all, subset = nCount_Xenium > 10 & cell_area >30)

all <- JoinLayers(all)
all[["Xenium"]] <- split(all[["Xenium"]], f = all$geo_accession)
all <- NormalizeData(all, verbose=F)
all <- FindVariableFeatures(all, nfeatures = 300, verbose=F)

all <- ScaleData(all, verbose=F)
all <- RunPCA(all, verbose=F)

all <- IntegrateLayers(object = all, method = RPCAIntegration, orig.reduction = "pca", new.reduction = "integrated.rpca",verbose = T)

all <- JoinLayers(all)
all<- RunUMAP(all, reduction = "integrated.rpca", dims=1:30)

all <- FindNeighbors(all, dims=1:30, reduction = "integrated.rpca")
all <- FindClusters(all, resolution = 0.4)

UMAPPlot(all, label=T)+FeaturePlot(all, "CD8A")
saveRDS(all, "../data/GSE307895_processed/all_int.rds")
```


```{r}
DotPlot(all, c( "CD8A", "EPCAM"))+coord_flip()+RotatedAxis()
```


```{r}
FetchData(all, c("geo_accession","nugentscore" ,"nugent", "leuteul", "status", "ident", "cell_area")) %>% group_by(geo_accession) %>% mutate(cells = n(), Treg = sum(ident == 36), Treg_per = 100*(Treg/cells), Treg_dens = Treg/sum(cell_area)) %>% 
  summarise(Treg_per = mean(Treg_per), 
            Treg_dens = mean(Treg_dens), 
            status = unique(status), 
            nugentscore = unique(nugentscore), 
            leuteal = unique(leuteul),
            nugent = unique(nugent)) %>% mutate(nugent = ifelse(nugent == "bacterial vaginosis", "BV", "Normal"), nugent = factor(nugent, levels=c("Normal", "BV"))) %>% 
  #ggplot(aes(Treg_per, as.numeric(nugentscore)))+geom_smooth(method='lm')
  #ggplot(aes(status, Treg_per))+geom_boxplot()
  ggplot(aes(status, Treg_per, fill = nugent))+geom_boxplot()+ggtitle("Treg (36)")

FetchData(all, c("geo_accession","nugentscore" ,"nugent", "leuteul", "status", "ident", "cell_area")) %>% group_by(geo_accession) %>% mutate(cells = n(), Treg = sum(ident == 29), Treg_per = 100*(Treg/cells), Treg_dens = Treg/sum(cell_area)) %>% 
  summarise(Treg_per = mean(Treg_per), 
            Treg_dens = mean(Treg_dens), 
            status = unique(status), 
            nugentscore = unique(nugentscore), 
            leuteal = unique(leuteul),
            nugent = unique(nugent)) %>% mutate(nugent = ifelse(nugent == "bacterial vaginosis", "BV", "Normal"), nugent = factor(nugent, levels=c("Normal", "BV"))) %>% 
  #ggplot(aes(Treg_per, as.numeric(nugentscore)))+geom_smooth(method='lm')
  #ggplot(aes(status, Treg_per))+geom_boxplot()
  ggplot(aes(status, Treg_per, fill = nugent))+geom_boxplot()+ggtitle("Treg (29)")

FetchData(all, c("geo_accession","nugentscore" ,"nugent", "leuteul", "status", "ident", "cell_area")) %>% group_by(geo_accession) %>% mutate(cells = n(), Treg = sum(ident == 16), Treg_per = 100*(Treg/cells), Treg_dens = Treg/sum(cell_area)) %>% 
  summarise(Treg_per = mean(Treg_per), 
            Treg_dens = mean(Treg_dens), 
            status = unique(status), 
            nugentscore = unique(nugentscore), 
            leuteal = unique(leuteul),
            nugent = unique(nugent)) %>% mutate(nugent = ifelse(nugent == "bacterial vaginosis", "BV", "Normal"), nugent = factor(nugent, levels=c("Normal", "BV"))) %>% 
  #ggplot(aes(Treg_per, as.numeric(nugentscore)))+geom_smooth(method='lm')
  #ggplot(aes(status, Treg_per))+geom_boxplot()
  ggplot(aes(status, Treg_per, fill = nugent))+geom_boxplot()+ggtitle("Treg (16)")

FetchData(all, c("geo_accession","nugentscore" ,"nugent", "leuteul", "status", "ident", "cell_area")) %>% group_by(geo_accession) %>% mutate(cells = n(), Treg = sum(ident == 36), Treg_per = 100*(Treg/cells), Treg_dens = Treg/sum(cell_area)) %>% 
  summarise(Treg_per = mean(Treg_per), 
            Treg_dens = mean(Treg_dens), 
            status = unique(status), 
            nugentscore = unique(nugentscore), 
            leuteal = unique(leuteul),
            nugent = unique(nugent)) %>% mutate(nugent = ifelse(nugent == "bacterial vaginosis", "BV", "Normal"), nugent = factor(nugent, levels=c("Normal", "BV"))) %>% 
  #ggplot(aes(Treg_per, as.numeric(nugentscore)))+geom_smooth(method='lm')
  #ggplot(aes(status, Treg_per))+geom_boxplot()
  ggplot(aes(status, Treg_dens, fill = nugent))+geom_boxplot()+theme(axis.text.y = element_blank())+ggtitle("Treg (36)")

FetchData(all, c("geo_accession","nugentscore" ,"nugent", "leuteul", "status", "ident", "cell_area")) %>% group_by(geo_accession) %>% mutate(cells = n(), Treg = sum(ident == 29), Treg_per = 100*(Treg/cells), Treg_dens = Treg/sum(cell_area)) %>% 
  summarise(Treg_per = mean(Treg_per), 
            Treg_dens = mean(Treg_dens), 
            status = unique(status), 
            nugentscore = unique(nugentscore), 
            leuteal = unique(leuteul),
            nugent = unique(nugent)) %>% mutate(nugent = ifelse(nugent == "bacterial vaginosis", "BV", "Normal"), nugent = factor(nugent, levels=c("Normal", "BV"))) %>% 
  #ggplot(aes(Treg_per, as.numeric(nugentscore)))+geom_smooth(method='lm')
  #ggplot(aes(status, Treg_per))+geom_boxplot()
  ggplot(aes(status, Treg_dens, fill = nugent))+geom_boxplot()+theme(axis.text.y = element_blank())+ggtitle("Treg (29)")

FetchData(all, c("geo_accession","nugentscore" ,"nugent", "leuteul", "status", "ident", "cell_area")) %>% group_by(geo_accession) %>% mutate(cells = n(), Treg = sum(ident == 16), Treg_per = 100*(Treg/cells), Treg_dens = Treg/sum(cell_area)) %>% 
  summarise(Treg_per = mean(Treg_per), 
            Treg_dens = mean(Treg_dens), 
            status = unique(status), 
            nugentscore = unique(nugentscore), 
            leuteal = unique(leuteul),
            nugent = unique(nugent)) %>% mutate(nugent = ifelse(nugent == "bacterial vaginosis", "BV", "Normal"), nugent = factor(nugent, levels=c("Normal", "BV"))) %>% 
  #ggplot(aes(Treg_per, as.numeric(nugentscore)))+geom_smooth(method='lm')
  #ggplot(aes(status, Treg_per))+geom_boxplot()
  ggplot(aes(status, Treg_dens, fill = nugent))+geom_boxplot()+theme(axis.text.y = element_blank())+ggtitle("Treg (16)")

FetchData(all, c("geo_accession","nugentscore" ,"nugent", "leuteul", "status", "ident", "cell_area")) %>% group_by(geo_accession) %>% mutate(cells = n(), Treg = sum(ident == 3), Treg_per = 100*(Treg/cells), Treg_dens = Treg/sum(cell_area)) %>% 
  summarise(Treg_per = mean(Treg_per), 
            Treg_dens = mean(Treg_dens), 
            status = unique(status), 
            nugentscore = unique(nugentscore), 
            leuteal = unique(leuteul),
            nugent = unique(nugent)) %>% mutate(nugent = ifelse(nugent == "bacterial vaginosis", "BV", "Normal"), nugent = factor(nugent, levels=c("Normal", "BV"))) %>% 
  #ggplot(aes(Treg_per, as.numeric(nugentscore)))+geom_smooth(method='lm')
  #ggplot(aes(status, Treg_per))+geom_boxplot()
  ggplot(aes(status, Treg_dens, fill=nugent))+geom_boxplot()+theme(axis.text.y = element_blank())+ggtitle("CD8 (3)")
```

## lisaClust

```{r}
all2 = all %>% FetchData(c("x_centroid", "y_centroid", "orig.ident", "ident")) %>%
  rename(all_of(c(x = "x_centroid", y = "y_centroid", imageID = "orig.ident", cellType = "ident"))) %>%
  lisaClust(., k=5)

all$region = all2$region
all %>% FetchData(c("x_centroid", "y_centroid", "region", "geo_accession", "COL17A1")) %>%
  filter(geo_accession == "GSM9233671") %>%
  ggplot(aes(x_centroid, y_centroid, color = region)) +geom_point()


all %>% FetchData(c("x_centroid", "y_centroid", "region", "geo_accession", "status")) %>%
  group_by(geo_accession) %>%
  mutate(total_cells = n()) %>%
  group_by(geo_accession, region) %>% 
  mutate(region_per = 100*(n()/total_cells)) %>%
  summarise(region_per = mean(region_per), status=unique(status)) %>%
    group_by(status, region) %>%
    summarise(region_per = mean(region_per)) %>%
  ggplot(aes(status, region_per, fill=region))+geom_col()+
  #facet_wrap(~status, scales="free")+
  RotatedAxis()
```

## spatially variable genes

```{r}
library(nnSVG)
m <- all@assays$Xenium@layers$data[,all$geo_accession == "GSM9233671"]
rownames(m) <- rownames(all); colnames(m) <- colnames(all)[all$geo_accession == "GSM9233671"]
all2 <- nnSVG(input = m, 
      spatial_coords = data.frame(`x` = all$x_centroid[all$geo_accession == "GSM9233671"], `y` = all$y_centroid[all$geo_accession == "GSM9233671"], row.names = colnames(all)[all$geo_accession == "GSM9233671"]))
```






### Add Compartment metadata

```{r}
tissue <- raster::raster("../data/test_mask.tif")# test region of the first sample acquired above. To now go make a 

coord = FetchData(dat, c("x_centroid", "y_centroid"))/0.2125
coords <- coord
coordinates(coord) <- ~x_centroid+y_centroid

dat <- AddMetaData(dat, 
                   metadata=na.omit(data.frame(tissue = raster::extract(tissue, coord), 
                   row.names = rownames(coords))))
FeaturePlot(dat,"tissue", reduction = "coordinates")
```

```{r}

```


```{r}

```

### Data Summary

```{r data-summary}
cat("Cells:      ", ncol(xen), "\n")
cat("Genes:      ", nrow(xen), "\n")
cat("Median transcripts per cell:", median(xen$nCount_Xenium), "\n")
```


## Quality Control

### QC Metric Distributions

```{r qc-metrics}
# Violin plots of key QC metrics
VlnPlot(
  xen,
  features  = c("nCount_Xenium", "nFeature_Xenium"),
  ncol      = 2,
  pt.size   = 0
)
```

### Spatial QC Overview

```{r qc-spatial}
# Spatial scatter of transcript counts
ImageFeaturePlot(xen, features = "nCount_Xenium")
```

### Cell Area Distribution

```{r qc-area}
# Requires area_um2 in metadata
# ggplot(xen@meta.data, aes(x = area_um2)) +
#   geom_histogram(bins = 100, fill = "steelblue") +
#   geom_vline(xintercept = c(params$min_area_um2, params$max_area_um2),
#              linetype = "dashed", colour = "red") +
#   labs(x = "Cell area (µm²)", y = "Count", title = "Cell area distribution")
```

### Filtering

```{r qc-filter}
cells_before <- ncol(xen)

xen <- subset(
  xen,
  nCount_Xenium >= params$min_transcripts
  # & area_um2 >= params$min_area_um2
  # & area_um2 <= params$max_area_um2
)

cells_after <- ncol(xen)

cat(sprintf("Cells before filtering: %d\nCells after filtering:  %d\nRemoved: %d (%.1f%%)\n",
  cells_before, cells_after,
  cells_before - cells_after,
  100 * (cells_before - cells_after) / cells_before))
```

---

## Normalisation & Scaling

Xenium 5K panels are targeted assays with a limited, curated gene set. Based on
benchmarking evidence (e.g. Dannenfelser et al. 2024, *Nature Methods*), standard
log-normalisation followed by scaling outperforms SCTransform for targeted spatial
panels and is the recommended approach here.

```{r normalise}
xen <- NormalizeData(xen, normalization.method = "LogNormalize", scale.factor = 10000)
xen <- ScaleData(xen, features = rownames(xen))
```

---

## Dimensionality Reduction

### PCA

```{r pca}
xen <- RunPCA(xen, npcs = params$n_pcs, verbose = FALSE)

ElbowPlot(xen, ndims = params$n_pcs) +
  ggtitle("Elbow plot — select number of PCs")
```

### UMAP

```{r umap}
xen <- RunUMAP(xen, dims = 1:params$n_pcs, verbose = FALSE)

DimPlot(xen, reduction = "umap", label = FALSE) +
  ggtitle("UMAP — all cells")
```

### t-SNE

```{r tsne}
xen <- RunTSNE(xen, dims = 1:params$n_pcs, verbose = FALSE)

DimPlot(xen, reduction = "tsne", label = FALSE) +
  ggtitle("t-SNE — all cells")
```

---

## Clustering

```{r cluster}
xen <- FindNeighbors(xen, dims = 1:params$n_pcs, k.param = params$n_neighbours)
xen <- FindClusters(xen, resolution = params$resolution)

cat("Clusters identified:", nlevels(xen$seurat_clusters), "\n")
table(xen$seurat_clusters)
```

### Cluster Visualisation

```{r cluster-plots}
p_umap  <- DimPlot(xen, reduction = "umap",  label = TRUE, repel = TRUE) + NoLegend()
p_tsne  <- DimPlot(xen, reduction = "tsne",  label = TRUE, repel = TRUE) + NoLegend()
p_pca   <- DimPlot(xen, reduction = "pca",   label = TRUE, repel = TRUE) + NoLegend()
p_space <- ImageDimPlot(xen, fov = "fov", cols = "polychrome", size = 0.5)

(p_umap | p_tsne) / (p_pca | p_space)
```

### Cluster Size Summary

```{r cluster-summary}
xen@meta.data %>%
  count(seurat_clusters, name = "n_cells") %>%
  mutate(pct = round(100 * n_cells / sum(n_cells), 1)) %>%
  kable(caption = "Cells per cluster") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

---

## Differential Expression

### Cluster Marker Discovery

```{r markers, cache=TRUE}
all_markers <- FindAllMarkers(
  xen,
  only.pos      = TRUE,
  min.pct       = 0.1,
  logfc.threshold = 0.25,
  test.use      = "wilcox"
)

# Save full marker table
write_csv(all_markers,
  file.path(params$out_dir, "tables", "all_cluster_markers.csv"))

# Preview top 5 per cluster
all_markers %>%
  group_by(cluster) %>%
  slice_max(avg_log2FC, n = 5) %>%
  kable(caption = "Top 5 markers per cluster (avg_log2FC)") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

### Top Marker Heatmap

```{r de-heatmap}
top_markers <- all_markers %>%
  group_by(cluster) %>%
  slice_max(avg_log2FC, n = 5) %>%
  pull(gene)

DoHeatmap(
  xen,
  features = unique(top_markers),
  group.by = "seurat_clusters"
) + NoLegend()
```

### Spatial Marker Overlay

```{r spatial-markers}
# Plot top marker for each of the first 4 clusters
top4 <- all_markers %>%
  group_by(cluster) %>%
  slice_max(avg_log2FC, n = 1) %>%
  ungroup() %>%
  slice(1:4) %>%
  pull(gene)

ImageFeaturePlot(xen, features = top4, ncol = 2)
```

### Excel Output — All Markers

```{r de-excel}
wb <- createWorkbook()

for (cl in sort(unique(all_markers$cluster))) {
  addWorksheet(wb, sheetName = paste0("Cluster_", cl))
  writeData(wb,
    sheet = paste0("Cluster_", cl),
    x     = filter(all_markers, cluster == cl))
}

saveWorkbook(wb,
  file      = file.path(params$out_dir, "tables", "cluster_markers.xlsx"),
  overwrite = TRUE)

cat("Marker Excel written to:", file.path(params$out_dir, "tables", "cluster_markers.xlsx"))
```

---

## Pathway Analysis

```{r pathway-check, eval=params$run_pathway}
# This section runs only when params$run_pathway == TRUE
```

### Gene Set Enrichment per Cluster

```{r pathway-gsea, eval=params$run_pathway}
# Convert gene symbols to Entrez IDs and run enrichment per cluster.
# Swap org.Hs.eg.db -> org.Mm.eg.db for mouse data.

pathway_results <- lapply(sort(unique(all_markers$cluster)), function(cl) {

  genes <- all_markers %>%
    filter(cluster == cl, p_val_adj < 0.05) %>%
    pull(gene)

  entrez <- bitr(genes,
    fromType = "SYMBOL",
    toType   = "ENTREZID",
    OrgDb    = org.Hs.eg.db)$ENTREZID

  if (length(entrez) < 5) return(NULL)

  if (params$pathway_db == "GO") {
    enrichGO(gene = entrez, OrgDb = org.Hs.eg.db,
             ont = "BP", pAdjustMethod = "BH",
             pvalueCutoff = 0.05, readable = TRUE)
  } else if (params$pathway_db == "KEGG") {
    enrichKEGG(gene = entrez, organism = "hsa",
               pAdjustMethod = "BH", pvalueCutoff = 0.05)
  } else if (params$pathway_db == "Reactome") {
    # library(ReactomePA)
    # enrichPathway(gene = entrez, organism = "human", readable = TRUE)
    NULL
  }
})

names(pathway_results) <- paste0("Cluster_", sort(unique(all_markers$cluster)))
```

### Pathway Dot Plots

```{r pathway-plots, eval=params$run_pathway, fig.height=6}
for (nm in names(pathway_results)) {
  res <- pathway_results[[nm]]
  if (!is.null(res) && nrow(res) > 0) {
    print(
      dotplot(res, showCategory = 15, title = nm)
    )
  }
}
```

### Pathway Excel Output

```{r pathway-excel, eval=params$run_pathway}
wb_path <- createWorkbook()

for (nm in names(pathway_results)) {
  res <- pathway_results[[nm]]
  if (!is.null(res) && nrow(res) > 0) {
    addWorksheet(wb_path, sheetName = nm)
    writeData(wb_path, sheet = nm, x = as.data.frame(res))
  }
}

saveWorkbook(wb_path,
  file      = file.path(params$out_dir, "tables", "pathway_results.xlsx"),
  overwrite = TRUE)
```

---

## Standard Visualisation Package

### Embedding Plots

```{r viz-embeddings}
DimPlot(xen, reduction = "umap",  label = TRUE, repel = TRUE) + ggtitle("UMAP") |
DimPlot(xen, reduction = "tsne",  label = TRUE, repel = TRUE) + ggtitle("t-SNE") |
DimPlot(xen, reduction = "pca",   label = TRUE, repel = TRUE) + ggtitle("PCA")
```

### Spatial Cluster Overlay

```{r viz-spatial}
ImageDimPlot(xen, fov = "fov", cols = "polychrome", size = 0.5) +
  ggtitle("Spatial cluster map")
```

### Feature Plots — Selected Markers

```{r viz-features}
# Edit gene list as appropriate
genes_of_interest <- head(unique(top_markers), 9)

FeaturePlot(xen, features = genes_of_interest, ncol = 3,
            reduction = "umap", order = TRUE)
```

### Spatial Feature Plots — Selected Markers

```{r viz-spatial-features}
ImageFeaturePlot(xen, features = head(genes_of_interest, 4), ncol = 2)
```

### Violin Plots — Selected Markers

```{r viz-violin}
VlnPlot(xen, features = head(genes_of_interest, 6),
        group.by = "seurat_clusters", ncol = 3, pt.size = 0)
```

---

## Save Outputs

```{r save-outputs}
# Seurat object
saveRDS(xen, file.path(params$out_dir, paste0(params$sample_id, "_seurat.rds")))

# Cell metadata
write_csv(xen@meta.data,
  file.path(params$out_dir, "tables", "cell_metadata.csv"))

# Embeddings
lapply(c("umap", "tsne", "pca"), function(red) {
  emb <- as.data.frame(Embeddings(xen, reduction = red))
  emb$cell_id <- rownames(emb)
  write_csv(emb, file.path(params$out_dir, "tables",
    paste0(red, "_embeddings.csv")))
})

cat("All outputs written to:", params$out_dir)
```

---

# Session Information
```{r session-info}
sessionInfo()
```
